---
title: "GDP Data Analysis"
author: "Natalia Navarrete"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    code_folding: hide
---


# 1. Introduction


The GDP spreadsheet contains a list of tables related to different gross domestic product (gdp) and national income data.

##Structure of the Spreadsheet
Sheet title: Unique identifier for each table such as T10101.
Sheet rows: The description of thetable, which indicates the type of GDP-related data it contains.

##Categories of Data in Spreadsheet
The data is categorized into several major topics:
1. Overall GDP Trends ()
  - 
  
2.

3.

4.

5.

6.

7.

8.

The gdp database includes 
What this dataset is?
Purpose of my analysis?
How the document is structured.

GDP dataset.


The dataset includes:

Overall GDP indicators (T10101 - T10111)
Breakdown by product type (T10201 - T10206)
Breakdown by sector (T10301 - T10306)
Price indexes & inflation indicators (T10404, T10604, etc.)
National income and corporate data (T11000 - T11600)

# 2. Libraries
Explain which libraries you're using and why?

```{r libraries}

library(readxl)
library(dplyr)
library(here)
library(purrr)
library(tidyr)
library(ggplot2)
library(lubridate)
library(stringr)
library(tibble)
```

# 3. Data Import
Explain the data source, the file structure, and how you are importing it. 
```{r GDP Data}
gdp_path <- file.path("data", "macroeconomic", "gdp", "gdp_domestic_product_and_income.xlsx")

reshape_gdp <- function(sheet_name, skip_rows = 7) {
  # Determine frequency from sheet name:
  frequency <- ifelse(grepl("-A$", sheet_name), "Annual",
                      ifelse(grepl("-Q$", sheet_name), "Quarterly", NA))
  
  # Get the table code by removing the trailing "-A" or "-Q"
  table_code <- sub("-[AQ]$", "", sheet_name)
  
  # Read the sheet; skip the first 7 rows so that row 8 (the header) is read as data
  raw_data <- read_excel(gdp_path, sheet = sheet_name, skip = skip_rows, col_names = FALSE)
  
  # Extract the header row (row 8 in Excel becomes row 1 here)
  header <- as.character(raw_data[1, ])
  header[1] <- "Line"        
  header[2] <- "Indicator"   
  header[3] <- "Code"        
  colnames(raw_data) <- header
  
  # Remove the header row from the data
  data_clean <- raw_data[-1, ]
  
  # Pivot the data from wide to long format.
  tidy_data <- data_clean %>%
    pivot_longer(
      cols = -c(Line, Indicator, Code),
      names_to = "Date",
      values_to = "Value"
    )
  
  # Ensure the Date column is cleaned up
  tidy_data <- tidy_data %>%
    mutate(Date = str_trim(Date))  # Remove spaces
  
  # Now, filter the rows so that only valid date headers are kept.
  if (frequency == "Annual") {
    tidy_data <- tidy_data %>% filter(grepl("^[0-9]{4}$", Date))  # Keep YYYY format
  } else if (frequency == "Quarterly") {
    tidy_data <- tidy_data %>% filter(grepl("^[0-9]{4}Q[1-4]$", Date))  # Keep YYYYQX format
  }
  
  # Convert Date and Value columns appropriately
  tidy_data <- tidy_data %>%
    mutate(
      Date = case_when(
        frequency == "Annual" ~ as.character(Date),  # Store years as strings
        frequency == "Quarterly" ~ paste0(substr(Date, 1, 4), "-Q", substr(Date, 6, 6)),  # Store "YYYY-QX"
        TRUE ~ NA_character_
      ),
      Value = suppressWarnings(as.numeric(Value)),  # Convert Value to numeric
      Frequency = frequency,
      Sheet = sheet_name,
      TableCode = table_code
    )
  
  # Debugging: Print detected unique dates to verify correctness
  print(paste("Sheet:", sheet_name, "- Unique Dates:", paste(unique(tidy_data$Date)[1:10], collapse = ", ")))
  
  return(tidy_data)
}
```

# 4. Data Wrangling & Overview
Show how to filter or create subsets.
Summary statistics
Explain meaning of these subsets.

```{r filtering data,}
# Get all sheet names and remove non-data sheets (like "Contents")
all_sheets <- excel_sheets(gdp_path)
data_sheets <- all_sheets[all_sheets != "Contents"]

# Load all sheets into a list
gdp_list <- map(data_sheets, ~ reshape_gdp(.x, skip_rows = 7))

# Combine all sheets into one data frame
gdp_all <- bind_rows(gdp_list)

# Now you can filter for the type of data you need:
gdp_annual <- gdp_all %>% filter(Frequency == "Annual")
gdp_quarterly <- gdp_all %>% filter(Frequency == "Quarterly")
gdp_T10105   <- gdp_all %>% filter(TableCode == "T10105")
```

# 5. Exploratory Data Analysis
Use subsections to explore specific topics.
5.1 Overall GDP Trends
5.2 Sector Breakdown
5.3 Price Indexes & Inflation

```{r exploring data}
# Check the results
print(head(gdp_all))
print(head(gdp_annual))
print(head(gdp_quarterly))
print(head(gdp_T10105))

```


# 6. Basic Statistical / Time Series
You can include any statistical tests
6.1 Stationarity Tests
6.2 Simple Forecasting


Time Series Analysis (GDP Over Time)
Track GDP trends using "T10105" (Gross Domestic Product).
Compare annual vs. quarterly GDP (T10105-A vs. T10105-Q).
Calculate growth rates (T10101 - Percent Change in GDP).


library(ggplot2)
gdp_annual <- gdp_data_list[["T10105-A"]]
colnames(gdp_annual) <- make.names(colnames(gdp_annual))
gdp_annual <- gdp_annual %>% rename(Date = 1, GDP = 2)

ggplot(gdp_annual, aes(x = Date, y = GDP)) +
  geom_line() +
  labs(title = "Annual GDP Trends", x = "Year", y = "GDP (Billions)")


Sector Contribution to GDP
Identify the fastest-growing sectors using "T10305" (Gross Value Added by Sector).
Compare GDP share by sector over time.

sector_gdp <- gdp_data_list[["T10305-A"]]
ggplot(sector_gdp, aes(x = Date, y = Manufacturing)) +
  geom_line() +
  labs(title = "Manufacturing Sector GDP Contribution", x = "Year", y = "GDP (Billions)")


Inflation & Price Indexes
Use "T10104" (Price Indexes for GDP) and "T10604" (Price Indexes for Purchases).
Compare real vs. nominal GDP using "T10106" (Chained Dollars).

price_index <- gdp_data_list[["T10104-A"]]
ggplot(price_index, aes(x = Date, y = PriceIndex)) +
  geom_line() +
  labs(title = "GDP Price Index Over Time", x = "Year", y = "Index Value")


Recession Analysis
Look for negative GDP growth in "T10101" (Percent Change in GDP).
Identify declining economic activity using "T10405" (Relation of GDP to Purchases).

gdp_growth <- gdp_data_list[["T10101-A"]]
gdp_growth <- gdp_growth %>% mutate(Recession = ifelse(GrowthRate < 0, "Recession", "Growth"))

ggplot(gdp_growth, aes(x = Date, y = GrowthRate, color = Recession)) +
  geom_line() +
  labs(title = "GDP Growth Rate and Recession Periods", x = "Year", y = "Growth Rate (%)")


Predicting Future GDP
Apply time series forecasting (auto.arima()) on "T10105".

library(forecast)
gdp_ts <- ts(gdp_annual$GDP, start = min(gdp_annual$Date), frequency = 1)
model <- auto.arima(gdp_ts)
forecasted_gdp <- forecast(model, h = 5)
autoplot(forecasted_gdp)

# 7 Conclusions and Next Steps
Summarize key insights from your analysis:
Major trends or shifts in GDP.
Which sectors drive GDP growth.
Relationship between price indexes and nominal/real GDP.
Any preliminary forecast results.
Suggest next steps or more advanced analyses (e.g., advanced econometric modeling, comparison with external data like unemployment rates or interest rates).

# 8 Appendix
Include any additional tables, charts, or diagnostic checks you want to show but keep out of the main narrative.
You might also include session info for reproducibility.

